Type theory in type theory with single substitutions

### 2024-09-18 types_2024_postproc/p.tex
tuff

In our single substitution calculus, we have equation $[\p][^+]$
stated both for types and terms (we know that the left one is about
types because of our convention that type metavariables are uppercase
Latin letters and term metavariables are lowercase):
\[
B[\p][\gamma^+] = B[\gamma][\p] \hspace{4em} b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]
\]
We even need the first equation to typecheck the second one. We made
the dependency explicit by adding a subscript of the equality in the
equation for terms (in Agda, this dependency has to be made
explicit). An alternative presentation of the second equation without
requiring the first one is the following:
\[
[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p]
\]
That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
we have that for any $b$ we have the equation (suitably over the input
equation). Thus we add a conditional equation.

@@

\begin{definition}[Minimised single substitution calculus with $\Pi$ and $\U$]\label{def:min}
  This section defined the minimised first-order version of the SOGAT
  (\ref{eq:tt}), relying essentially on the presernce of universes and
  $\Pi$ types. For reference, list all the rules of the minimised
  calculus in Appendix \ref{app:minimisation}.
\end{definition}
\begin{theorem}
  The GATs of Definition \ref{def:min} and Definition \ref{def:tt} are
  isomorphic, in particular, all equations are interderivable.
\end{theorem}
\begin{proof}
  Clear from the construction in this section.
\end{proof}

@@

  (\gamma\circ\p,\q)$ and $\langle a\rangle := (\id,a)$.

@@

  with $\un^k : \Tm\,(\Lift^k\,A)\cong\Tm\,A : \mk^k$. The category

@@

  $\Con$, so $\Delta$ can mean $\Delta_{.1}$ or $\Delta_{.2}$
  depending on the context, and we used the truncating subtraction of

@@

  appropriate (un)lifting. The category laws hold. The empty context is

@@

    & \Ty\,\Gamma\,i && := \Tm\,\diamond\,(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)\,(\U\,i)}) \\

@@

  To make the notation readable, we will not write the lifting
  decorations from now on, and we repeat the previous definitions
  without mentioning universe levels, but all of what follows is
  definable properly with lifting.

@@

it and comparing the result with the original SSC does not provide an

@@

Defining a useable syntax of type theory internally to homotopy type
theory (HoTT \cite{HoTTbook}) is an open problem: the (h)set-truncated
syntax can be formalised as a QIIT, but then we cannot even define the
$\Type$-interpretation, that is, a function from the syntax which
interprets closed types by the metatheoretic universe $\Type$ (rather
than an inductive recursive universe as in Subsection
\ref{sec:standard} or \cite[Section
  4]{DBLP:conf/popl/AltenkirchK16}). In two-level type theory the
syntax of type theory can be defined using infinitely many coherene
rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
$\Sub$, we have the associativity of composition of substitutions
equation. If we truncate our syntax to be a set, then this is
enough. If we don't truncate, there are two different ways of proving
$((\gamma\circ\delta)\circ\theta)\circ\xi =
\gamma\circ(\delta\circ(\theta\circ\xi))$, and nothing forces these to
be the same, so in such a (wild) syntax substitutions don't form a
set, hence they don't have decidable equality \cite{hedberg}. The
solution is to add the pentagon law, but then this introduces new
higher equalities, which require more coherence laws, so in the end we
have an infinite sequence of higher coherence laws that need to be
part of the syntax. Our single substitution calculus does not have
composition for $\Sub$, so there is no need for higher coherence laws
like associativity. However as it currently stands, our syntax is not
coherent, it needs set-truncation when working in HoTT. For example,
there are two different ways to prove the following equation:
\begin{alignat*}{10}
  & (\cd\,\U)[\p][\langle a\rangle] \overset{[\p][\langle\rangle]}{=} \cd\,\U \\
  & (\cd\,\U)[\p][\langle a\rangle] \overset{\cd[]}{=} \cd\,(\U[\p][\langle a\rangle]) \overset{\U[]}{=} \cd\,\U
\end{alignat*}
In a coherent version of our syntax we either need to state that these
two proofs are equal, or we have to remove one of them. We can remove
the first one by limiting $[\p][\langle\rangle]$ to only work on
variables (this needs the introduction of a new sort for
variables). If we do this, the minimisation of Section
\ref{sec:minimisation} does not work anymore, so we need the extra
four equations for types ($[\p][^+]$, $[\p][\langle\rangle]$,
$[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$). These equations on
types make the syntax non-coherent again. A possible way out of this
would be to make the equations that depend on one of these for
equations conditional. But then the methods of Section
\ref{sec:admissible} don't work anymore for re-proving the needed
admissible equations. Also, if only have the rule $[\p][^+]$ only for
variables, then we cannot redo the termification of Section
\ref{sec:cwf}, in particular it is not clear how to prove equation
(\ref{eq:functor}).

In this paper we unfolded a particular example of the SOGAT $\ra$
single substitution GAT translation
\cite{DBLP:conf/fscd/KaposiX24}. We showed that for this theory, the
syntax is isomorphic to the parallel syntax, and we conjecture that
our method works for any SOGAT. For our particular SOGAT, the single
substitution syntax can be simplified, and we even have that any
single substitution model is equivalent to a parallel model in a weak
sense. This shows that if we have enough type formers, the
substitution calculus can be very weak, and still enough to bootstrap
the usual theory. It needs further investigation whether we can remove
even more structural rules, and still obtain isomorphic syntaxes. We
would also like to understand what a good notion of weak equivalence
is for single substitution models.

>>>>

In our single substitution calculus, we have equation $[\p][^+]$
stated both for types and terms (we know that the left one is about
types because of our convention that type metavariables are uppercase
Latin letters and term metavariables are lowercase):
\[
B[\p][\gamma^+] = B[\gamma][\p] \hspace{4em} b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]
\]
We even need the first equation to typecheck the second one. We made
the dependency explicit by adding a subscript of the equality in the
equation for terms (in Agda, this dependency has to be made
explicit). An alternative presentation of the second equation without
requiring the first one is the following:
\[
[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p]
\]
That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
we have that for any $b$ we have the equation (suitably over the input
equation). Thus this is a conditional equation.

@@

\begin{definition}[Minimised single substitution calculus with $\Pi$ and $\U$]\label{def:min}
  This section defined the minimised first-order version of the SOGAT
  (\ref{eq:tt}), relying essentially on the presence of universes and
  $\Pi$ types. For reference, we list all the rules of the minimised
  calculus in Appendix \ref{app:minimisation}.
\end{definition}
\begin{theorem}
  The GATs of Definition \ref{def:min} and Definition \ref{def:tt} are
  isomorphic, in particular, all equations are interderivable.
\end{theorem}
\begin{proof}
  Clear from the construction in this section.
\end{proof}

@@

  (\gamma\circ\p,\q)$ and $\langle a\rangle := (\id,a)$. All equations are derivable.

@@

  with $\un^k : \Tm\,(\Lift^k\,A)$ $\cong\Tm\,A : \mk^k$. The category % hack

@@

  $\Con$, so $\Delta$ can mean $\Delta_{.1}$ or $\Delta_{.2}$, and we used the truncating subtraction of

@@

  appropriate (un)liftings. The category laws hold. The empty context is

@@

    & \Ty\,\Gamma\,i && := \Tm\,\diamond\,\big(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)}\,(\U\,i)\big) \\

@@

  To make the notation readable, from now on, we will not write the
  lifting decorations or universe from now on. We repeat the previous
  definitions again without writing decorations, but all of what
  follows can be defined properly.

@@

it and comparing the result with the original SSC-model does not provide an

@@

Defining a useable syntax of type theory internally to homotopy type
theory (HoTT \cite{HoTTbook}) is an open problem \cite{mike,nicolai}: the (h)set-truncated
syntax can be formalised as a QIIT, but then we cannot even define the
$\Type$-interpretation, that is, a function from the syntax which
interprets closed types in the metatheoretic universe $\Type$ (rather
than an inductive recursive universe as in Subsection
\ref{sec:standard} or \cite[Section
  4]{DBLP:conf/popl/AltenkirchK16}). In two-level type theory the
syntax of type theory can be defined using infinitely many coherene
rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
$\Sub$, we have an equation in the syntax on the associativity of composition of substitutions.
If we truncate our syntax to be a set, then this is
enough. If we don't truncate, there are two different ways of proving
$((\gamma\circ\delta)\circ\theta)\circ\xi =
\gamma\circ(\delta\circ(\theta\circ\xi))$, and nothing forces these to
be the same, so in such a (wild) syntax substitutions don't form a
set, hence they don't have decidable equality \cite{hedberg}. The
solution is to add the pentagon law, but then this introduces new
higher equalities, which require more coherence laws, so in the end we
have an infinite sequence of higher coherence laws that need to be
part of the syntax. Our single substitution calculus does not have
composition for $\Sub$, so there is no need for higher coherence laws
like associativity. However as it currently stands, our syntax is not
coherent, it needs set-truncation when working in HoTT. For example,
there are two different ways to prove the following equation:
\begin{alignat*}{10}
  & (\cd\,\U)[\p][\langle a\rangle] \overset{[\p][\langle\rangle]}{=} \cd\,\U \\
  & (\cd\,\U)[\p][\langle a\rangle] \overset{\cd[]}{=} \cd\,(\U[\p][\langle a\rangle]) \overset{\U[]}{=} \cd\,\U
\end{alignat*}
In a coherent version of our syntax we either need to state that these
two proofs are equal, or we have to remove one of them. We can remove
the first one by limiting $[\p][\langle\rangle]$ to only apply to
variables (this needs the introduction of a new sort for
variables). If we do this, the minimisation of Section
\ref{sec:minimisation} does not work anymore, so we need the extra
four equations for types ($[\p][^+]$, $[\p][\langle\rangle]$,
$[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$). These equations on
types make the syntax non-coherent again. A possible way out of this
would be to make the equations that depend on one of these 
 conditional. But then the methods of Section
\ref{sec:admissible} don't work anymore for re-proving the needed
admissible equations. Also, if we only have the rule $[\p][^+]$ for
variables, we cannot redo the termification of Section
\ref{sec:cwf}, e.g.\ it is not clear how to prove equation
(\ref{eq:functor}).

In this paper we unfolded a particular example of the SOGAT $\ra$
single substitution GAT translation
\cite{DBLP:conf/fscd/KaposiX24}. We showed that for this theory, the
syntax is isomorphic to the parallel syntax, and we conjecture that
our method works for any SOGAT. For our particular SOGAT, we minimised the single
substitution syntax, and we even have that any
single substitution model is equivalent to a parallel model in a weak
sense. This shows that if we have enough type formers, the
substitution calculus can be very weak, and still enough to bootstrap
the usual theory. It needs further investigation whether we can remove
even more structural rules, and still obtain isomorphic syntaxes. We
would also like to understand what a good notion of weak equivalence
is for single substitution models.
